\chapter*{Acknowledgements}
%\addcontentsline{toc}{chapter}{Acknowledgements}

I would like to express my sincere gratitude to my supervisor, Assistant Professor Tan Yong Kiam, for his patient guidance, constant encouragement, and generous support throughout this project.
His advice shaped both the mathematical direction and the formal development of this work, and his detailed feedback repeatedly helped me sharpen the structure, rigor, and presentation of the report.
I have particularly benefited from his careful comments on how to connect the formal results and present the Isabelle/HOL development in a more precise and meaningful way.
His insight and high standards greatly improved the quality of both the formalization and the written exposition.


I am also grateful for the LAMR reading group based on \textit{Logic and Mechanized Reasoning}, which helped me build background in computational logic, mechanized reasoning, and formal proof methods that proved useful for this project.

My thanks also go to the Isabelle/HOL and Archive of Formal Proofs communities for providing the libraries, documentation, and prior formal developments that made this project possible.
Their work has created a rich and reusable formal ecosystem on which this report could build.

Finally, I would like to acknowledge the broader academic environment in which this project was carried out.
The opportunity to study the interaction between theoretical computer science, probability, and mechanized reasoning has made this project both challenging and deeply rewarding.


