\chapter{Proof Construction}
\label{chp:proof}

In this chapter, we construct a machine-checked proof of the Isolation Lemma in Isabelle/HOL and present the development in three layers (PP/SP/FP) to bridge informal intuition and formal verification.
Unlike pen-and-paper proofs, formal verification requires every mathematical step to be made fully explicit.
This includes not only the main reasoning steps, but also background obligations such as existential witnesses, measurability, independence, type correctness, and totality.

Our goal is to construct a fully formal proof (FP) of the Isolation Lemma in Isabelle/HOL.
However, it is neither practical nor desirable to translate a pen-and-paper (PP) proof directly into formal code.
Instead, we introduce an intermediate stage, which we refer to as the structured proof (SP) level.
This stage presents a structured decomposition of the argument, designed to align with the proof obligations of Isabelle/HOL (e.g.\ existence, measurability, and type constraints) while remaining human-readable.

Thus, each proof component in this chapter is presented in three layers:
\begin{itemize}
    \item \textbf{PP:} The intuitive pen-and-paper sketches from the literature.
    \item \textbf{SP:} A structured and detailed decomposition refined to match the proof obligations of Isabelle/HOL.
    \item \textbf{FP:} The corresponding formalization code in Isabelle/HOL.
\end{itemize}

We organize the whole proof as a sequence of structural reductions:
We begin by establishing the existence of minimizers and by reformulating its non-uniqueness in a constructive form.
Then we analyze the auxiliary quantity $\alpha(x)$, which plays a central role in isolating the dependence on individual weights.
This leads to a reduction of the uniqueness failure to an explicit collision event of the form $w(x) = \alpha(x)$.
Finally, we bound the probability of such events using the product probability space constructed in Chapter~\ref{chp:modeling}.

\section{Existence of Minimizers}
\label{sec:existence}

Before analyzing the uniqueness of minimizers, we must first ensure that the notion of a minimizer is well-defined.
In particular, for a given family of sets $\mathcal{F}$ and a weight function $w$, it is necessary to establish that at least one minimizer exists whenever $\mathcal{F}$ is non-empty.

\paragraph{PP.}
In informal proofs, this step is usually omitted.
Since weights are natural number and the family $\mathcal{F}$ is assumed to be non-empty, the existence of a minimizer is considered self-evident.

\paragraph{SP.}
In preparation for formalization, however, we must isolate this assumption and state it precisely, since later constructs (like $\alpha(x)$ or uniqueness predicates) depend on the existence of at least one such set.

\noindent
\textbf{Lemma (Existence of a Minimizer).}
\emph{If $\mathcal{F} \neq \emptyset$, then there exists a set
$A \in \mathcal{F}$ such that $A$ is a minimizer of $\mathcal{F}$ under
the weight function $w$.}

\paragraph{FP.}
This lemma is formalized as:

\begin{lstlisting}[language=Isabelle, caption={Existence of Minimizer}]
lemma exist_minimizer:
  assumes "F \neq {}"
  obtains A where "A \in F" "minimizer F w A"
\end{lstlisting}

The proof proceeds by constructing the minimal weight value using Isabelle's \texttt{LEAST} operator.
First, we extract a witness $S_0 \in \mathcal{F}$ to ensure the minimum is well-defined.
We then define the minimal weight $m$ as the least $n$ such that some $S \in \mathcal{F}$ satisfies $wt\ w\ S = n$.
A corresponding set $A$ is then chosen to realize this minimum.
Finally, we show that this $A$ has weight at most that any other set in $\mathcal{F}$, thus satisfying the \texttt{minimizer} predicate introduced in Chapter~\ref{chp:modeling}.

This lemma plays a foundational role: it ensures that future uses of the \texttt{LEAST} operator or comparisons among minimizers are grounded in an actual, verifiable witness.
By discharging this assumption with a lemma, subsequent lemmas can avoid repeatedly re-establishing this basic existence condition.


\section{Reformulating the Non-Uniqueness}
\label{sec:non-unique}

The Isolation Lemma asserts that with high probability, the weight function $w$ yields a unique minimizer in $\mathcal{F}$.
To prove this in a formal setting, we must first address the failure of uniqueness -- i.e., the possibility that multiple sets in $\mathcal{F}$ achieve the minimum weight.

\paragraph{PP.}
In pen-and-paper proofs, non-uniqueness is usually described in a purely intuitive way:
saying that a minimizer is "not unique" simply means that there exist at least two different sets in $\mathcal{F}$ that both attain the minimum weight.
This idea is often expressed informally by statements such as "there are two distinct minimizers", without any further logical unpacking.

\paragraph{SP.}
To make this intuition compatible with formal reasoning,we refine it into a positive existential formulation.
Instead of negating a uniqueness statement, we characterize failure of uniqueness by the explicit existence of two distinct witnesses that both satisfy the minimizer predicate.

Concretely, the minimizer of $\mathcal{F}$ under $w$ is not unique if and only if there exist two distinct sets $A \neq B$ in $\mathcal{F}$ such that both are minimizers.
In logical terms, this corresponds to replacing a negated unique-existence statement by an equivalent existential condition of the form
\[
\exists A\, B.\;
A \in \mathcal{F} \;\land\; B \in \mathcal{F} \;\land\;
\text{minimizer}(\mathcal{F}, w, A) \;\land\;
\text{minimizer}(\mathcal{F}, w, B) \;\land\;
A \neq B.
\]
This reformulation eliminates the outer negation and transforms the non-uniqueness into concrete objects that can be fixed in subsequent arguments.

Moreover, if $A$ and $B$ are both minimizers, then they must attain the same total weight.
This follows directly from the definition of the minimizer predicate:
each has weight at most that of any other set in $\mathcal{F}$, and hence in particular at most that of the other.
This elementary but explicit observation will be used later when comparing the contribution of individual elements in two different minimizers.

\noindent
\textbf{Lemma (Non-Uniqueness via Two Minimizers).}
\emph{Assume $\mathcal{F} \neq \emptyset$. Then the minimizer of $\mathcal{F}$ under $w$ is not unique if and only if there exist two distinct sets $A, B \in \mathcal{F}$ such that both $A$ and $B$ are minimizers; i.e., there exist $A \neq B$ in $\mathcal{F}$ with
\[
\forall S \in \mathcal{F},\; w(A) \le w(S) \;\land\; w(B) \le w(S),
\]
and in particular $w(A) = w(B)$.}

\paragraph{FP.}
These facts are formalized as:

\begin{lstlisting}[language=Isabelle, caption={Non-Uniqueness Characterization}]
lemma not_unique_iff_exist_two:
  assumes "F \neq {}"
  shows "\not has_unique_minimizer F w <-> 
        (EX A B. A \in F & B \in F &
                minimizer F w A & 
                minimizer F w B &
                A \neq B)"
\end{lstlisting}

\begin{lstlisting}[language=Isabelle, caption={Minimizers Have Equal Weight}]
lemma minimizers_same_weight:
  assumes "minimizer F w A" "minimizer F w B"
  shows "wt w A = wt w B"
\end{lstlisting}

This reformulation replaces a negated uniqueness predicate by a positive existential statement that provides concrete witnesses.
Once such witnesses $A$ and $B$ are fixed, later arguments can reason about their structural relationship, for example by analyzing elements in the symmetric difference $A \triangle B$.
This will be a key step in the collision characterization developed in subsequent sections.

As an immediate consequence of this characterization, if $A$ and $B$ are both minimizers, they must attain the same total weight.
This follows directly from the definition of the minimizer predicate and will be used implicitly in later comparisons between different minimizers.

\section{The Alpha Construction}
\label{sec:alpha}

A central ingredient in the proof of the Isolation Lemma is the auxiliary quantity $\alpha(x)$ associated with each element $x \in U$.
Intuitively, $\alpha(x)$ isolates the dependence of the total weight of a set on the individual weight $w(x)$, allowing us to reason about the effect of modifying a single coordinate.

Formally, $\alpha(x)$ is defined as the difference between two minimal weight values determined by the family $\mathcal{F}$.
The first term represents the minimum total weight among all sets in $\mathcal{F}$ that do not contain $x$.
The second term represents the minimum residual weight required to form a set in $\mathcal{F}$ that contains $x$, excluding the contribution of $x$ itself.
This definition was introduced by Mulmuley et al.\ and plays a crucial role in isolating single-coordinate events.

\medskip

In our formalization, $\alpha(x)$ is defined using Isabelle's \texttt{LEAST} operator, which selects the least natural number satisfying a given predicate.
We first define $\alpha(x)$ via \texttt{LEAST} to avoid committing to specific witnesses; the witness-based simplifications are derived later from the existence-of-minimizer lemmas.

\medskip

In the present chapter, we analyze its properties and show how the abstract \texttt{LEAST}-based definition can be simplified in subsequent arguments, once the existence of minimizers has been derived.

\subsection{Simplifying the Definition of \texorpdfstring{$\alpha(x)$}{alpha(x)}}

\paragraph{PP.}
In pen-and-paper proofs, the two minimization terms in the definition of $\alpha(x)$ are usually treated as concrete minimum values realized by suitable sets in $\mathcal{F}$.
Once a minimizer exists, the least total weight among sets not containing $x$ is simply the weight of a minimizer that avoids $x$.
Likewise, the least residual weight among sets containing $x$ is obtained by removing $x$ from a minimizer that contains it.
This allows the abstract \texttt{LEAST}-based definition of $\alpha(x)$ to be replaced by an ordinary arithmetic expression involving concrete minimizers.

\paragraph{SP.}
To make this intuition precise, we analyze the two \texttt{LEAST} expressions in the definition of $\alpha(x)$ separately.

First, suppose that $B$ is a minimizer of $\mathcal{F}$ under the weight function $w$, and that $x \notin B$.
Since $B \in \mathcal{F}$ and attains the minimum total weight among all sets in $\mathcal{F}$, it satisfies the defining predicate of the first \texttt{LEAST} expression.
Moreover, any other set $S \in \mathcal{F}$ with $x \notin S$ must have weight at least $w(B)$.
Hence, the minimum total weight among all sets in $\mathcal{F}$ that do not contain $x$ is attained by $B$, and the first \texttt{LEAST} expression can be shown equal to the concrete value $W(B)$.

Second, suppose that $A$ is a minimizer of $\mathcal{F}$ and that $x \in A$.
Removing $x$ from $A$ yields a residual set whose total weight equals $w(A) - w(x)$.
Since $A \in \mathcal{F}$ and attains the minimum total weight among all sets in $\mathcal{F}$, any other set $S \in \mathcal{F}$ containing $x$ must have residual weight at least $w(A \setminus \{x\})$.
Therefore, the least value selected by the second \texttt{LEAST} operator coincides with the concrete value $w(A \setminus \{x\})$.

Together, these two observations eliminate both instances of the \texttt{LEAST} operator from the definition of $\alpha(x)$ whenever suitable minimizers are available.
They reduce the abstract definition to a concrete arithmetic form that can be used directly in subsequent arguments.

\paragraph{FP.}
These simplifications are formalized by the following auxiliary lemmas in Isabelle/HOL:

\begin{lstlisting}[language=Isabelle, caption={First LEAST Eliminated by a Minimizer}]
lemma minimizer_B_eq_least_not_in:
  assumes "minimizer F w B" "x ~: B"
  shows "(LEAST n. EX S. S \in F & x ~: S & wt w S = n) 
         = wt w B"
\end{lstlisting}

\begin{lstlisting}[language=Isabelle, caption={Second LEAST Eliminated by a Minimizer}]
lemma minimizer_A_eq_least_in_residual:
  assumes "finite U" "F \subseteq Pow U" "minimizer F w A" "x \in A"
  shows "(LEAST n. EX S. S \in F & x \in S & wt w (S - {x}) = n) 
        = wt w (A - {x})"
\end{lstlisting}



These lemmas connect the abstract \texttt{LEAST} expressions in the definition of $\alpha(x)$ with concrete weights of minimizers.
In particular, they justify replacing both minimization terms by explicit arithmetic values whenever a minimizer avoiding $x$ or containing $x$ is available.

As a result, the auxiliary quantity $\alpha(x)$ can be rewritten in terms of ordinary arithmetic expressions, which will be essential for proving its structural properties and for carrying out the probabilistic analysis in later sections.

\subsection{Independence of \texorpdfstring{$\alpha(x)$}{alpha(x)} from Individual Weights}
\label{sec:alpha_independence}

\paragraph{PP.}
Intuitively, $\alpha(x)$ should not depend on the particular value of the single coordinate $w(x)$.
Changing the value of $w(x)$ shifts the total weight of every set containing $x$ by the same offset, while leaving the relative ordering of weights determined by the remaining elements unchanged.
Hence the two minima defining $\alpha(x)$ -- one over sets excluding $x$, and one over residual weights of sets including $x$ -- remain invariant under modifications of $w(x)$.

\paragraph{SP.}
We now formulate this independence property as a formal lemma, which will serve as a key technical interface for the probabilistic argument in the next section.

\noindent
\textbf{Lemma (Independence of $\alpha(x)$).}
\emph{Let $w$ and $w'$ be two weight functions such that $w(y) = w'(y)$ for all $y \in U \setminus \{x\}$.
Then $\alpha(x)$ computed with respect to $w$ is equal to $\alpha(x)$ computed with respect to $w'$.}

To prove this lemma, we compare the two definitions of $\alpha(x)$ under $w$ and $w'$.
The core observation is that for any set $S \subseteq U$ with $x \notin S$, the total weight $wt\ w\ S$ depends only on values of $w$ on $S \subseteq U \setminus \{x\}$, and therefore equals $wt\ w' \S$.
The same reasoning applies to residual sets of the form $S - \{x\}$, since $S - \{x\} \subseteq U \setminus \{x\}$.

Using this congruence of weights, we show that the predicates appearing in the two \texttt{LEAST} expressions defining $\alpha(x)$ are equivalent under $w$ and $w'$.
\[
\exists S\in F.\; x\notin S \land wt\ w\ S = n
\quad\Longleftrightarrow\quad
\exists S\in F.\; x\notin S \land wt\ w'\ S = n,
\]
and similarly for the residual term
\[
\exists S\in F.\; x\in S \land wt\ w\ (S-\{x\}) = n
\quad\Longleftrightarrow\quad
\exists S\in F.\; x\in S \land wt\ w'\ (S-\{x\}) = n.
\]

Therefore, the two \texttt{LEAST} values coincide under $w$ and $w'$.
Unfolding the definition of $\alpha(x)$ then yields the desired conclusion of the lemma, namely $\alpha_w(x) = \alpha_{w'}(x)$.

\paragraph{FP.}
This lemma is formalized as \texttt{alpha\_independent\_of\_coord} in Isabelle/HOL:

\begin{lstlisting}[language=Isabelle, caption={Independence of alpha(x) from the x-coordinate}]
lemma alpha_independent_of_coord:
  assumes "finite U" "F \subseteq Pow U" "x \in U"
  shows "ALL w w'. (ALL y. y \in U - {x} --> w y = w' y)
        --> alpha (w :: 'a => nat) F x = alpha w' F x"
\end{lstlisting}

This lemma is proved by establishing weight congruence on sets contained in $U - \{x\}$, transferring this congruence to the defining predicates of the two \texttt{LEAST} expressions in $\alpha(x)$, and finally unfolding \texttt{alpha\_def}.


\subsection{Collision characterization via \texorpdfstring{$\alpha(x)$}{alpha(x)}}
\label{sec:collision}

\paragraph{PP.}
We now establish the key structural link between the failure of uniqueness and the auxiliary quantity $\alpha(x)$.

Assume that the minimizer of $\mathcal{F}$ under $w$ is not unique.
Then there exist two distinct minimizers $A,B \in \mathcal{F}$.
Choose an element $x \in A \setminus B$.
Since $A$ and $B$ are minimizers, their total weights coincide:
\[
w(A) = w(B).
\]
Moreover, in the definition of $\alpha(x)$, the minimum over sets not containing $x$ is realized by $B$, while the minimum over residual weights of sets containing $x$ is realized by $A \setminus \{x\}$.
Hence,
\[
\alpha(x)
= w(B) - w(A \setminus \{x\})
= w(A) - \bigl(w(A) - w(x)\bigr)
= w(x).
\]
Therefore, non-uniqueness implies the existence of an element $x \in U$ such that $w(x) = \alpha(x)$.

\paragraph{SP.}

The pen-and-paper argument above already matches closely the structure of the formal proof.
The only additional ingredients required for formalization are the explicit side conditions ensuring that the two \texttt{LEAST} expression in the definitions of $\alpha(x)$ are well-defined and can be eliminated.

Concretely, we assume that the universe $U$ is finite and that the family $\mathcal{F}$ is a collection of subsets of $U$.
Under these assumptions, the two auxiliary lemmas established in Section~\ref{sec:alpha} apply:
the first shows that the \texttt{LEAST} term over sets not containing $x$ is realized by $B$ whenever $x \notin B$,
and the second shows that the residual \texttt{LEAST} term over sets containing $x$ is realized by $A \setminus \{x\}$ whenever $x \in A$.

Together with the lemma stating that any two minimizers have equal total weight, these results allow the formal proof to mirror the pen-and-paper derivation step by step.
No additional combinatorial insight is required at this stage; the formalization consists entirely in assembling previously established lemmas and discharging the necessary finiteness and membership conditions.

\paragraph{FP.}

This reasoning is captured in Isabelle/HOL by the following lemma:

\begin{lstlisting}[language=Isabelle, caption={Collision Witnessed by a Single Coordinate}]
lemma two_minimizers_alpha_eq:
  assumes "finite U" "F \subseteq Pow U"
          "A \in F" "B \in F"
          "minimizer F w A" "minimizer F w B"
          "x \in A - B"
  shows   "alpha w F x = w x"
\end{lstlisting}


The proof unfolds the definition of $\alpha(x)$ and eliminates both \texttt{LEAST} expressions using the auxiliary lemmas from Section~\ref{sec:alpha}.
The equality of weights $w(A)=w(B)$ follows from the fact that $A$ and $B$ are both minimizers.
Finally, the standard splitting identity $w(A) = w(A \setminus \{x\}) + w(x)$ is used to derive the desired conclusion $\alpha(x) = w(x)$.


\section{Probabilistic Analysis and Final Bound}
\label{sec:probability}

\paragraph{PP.}

We now complete the proof of the Isolation Lemma by bounding the probability that the minimizer of $\mathcal{F}$ is not unique under a random weight assignment.
Throughout this section, weights are sampled according to the product probability space constructed in Chapter~\ref{chp:modeling}.

By the collision characterization established in Section~\ref{sec:collision}, the event that the minimizer is not unique can be reduced to a union of single-coordinate events, following the counting strategy established in the literature \cite{spencer1995probabilistic}:
\[
\{\, \text{the minimizer is not unique} \,\}
\subseteq
\bigcup_{x \in U} \{\, w(x) = \alpha(x) \,\}.
\]
Thus, it suffices to bound the probability of each event $\{\, w(x) = \alpha(x) \,\}$.

Fix an arbitrary element $x \in U$.
By the independence property proved in Section~\ref{sec:alpha_independence}, the auxiliary quantity $\alpha(x)$ depends only on the weights assigned to elements in $U \setminus \{x\}$.
Consequently, once all other weights have been fixed, $\alpha(x)$ becomes a constant with respect to the remaining random choice of $w(x)$.

Under the product probability space, the random variable $w(x)$ is distributed uniformly over the finite set $\{1,\dots,N\}$ and is independent of all other coordinates.
Therefore, the probability that $w(x)$ attains any particular value is exactly $1/N$, and in particular
\[
\Pr\bigl[\, w(x) = \alpha(x) \,\bigr] \le \frac{1}{N}.
\]

Since the universe $U$ is finite, we may apply the union bound:
\[
\Pr\bigl[\, \text{the minimizer is not unique} \,\bigr]
\le \sum_{x \in U} \Pr\bigl[\, w(x) = \alpha(x) \,\bigr]
\le \frac{|U|}{N}.
\]
Equivalently, the probability that the minimizer is unique satisfies
\[
\Pr\bigl[\, \text{the minimizer is unique} \,\bigr]
\ge 1 - \frac{|U|}{N}.
\]
This inequality constitutes the statement of the Isolation Lemma.


\paragraph{SP.}

The formal proof follows the same high-level structure as the pen-and-paper argument, but it must expose the proof obligations that are implicit informally.

To obtain an explicit probability bound inside Isabelle/HOL, the event $\{\, w(x) = \alpha(x) \,\}$ is decomposed into a finite union over $k \in \{1,\dots,N\}$:
\[
\{\, w(x) = \alpha(x) \,\}
\;=\;
\bigcup_{k=1}^{N} \{\, \alpha(x) = k \;\wedge\; w(x) = k \,\}.
\]
This technical decomposition allows the application of independence at the level of random variables together with uniform point-probability bounds.

To obtain a multiplicative bound on each summand, the independence between $w(x)$ and $\alpha(x)$ must be established at the level of random variables.
This is achieved by exploiting the product structure of the underlying PMF.
The restriction of a sample function to $\{x\}$ and to $U \setminus \{x\}$ yields two independent components.
Since $w(x)$ depends only on the first component, while $\alpha(x)$ depends only on the second (by the independence-of-$\alpha$ lemma), their compositions are independent random variables.
This yields
\[
\Pr[\, \alpha(x) = k \wedge w(x) = k \,]
=
\Pr[\, \alpha(x) = k \,] \cdot \Pr[\, w(x) = k \,].
\]

Finally, the construction of $\alpha(x)$ via minimum operators over finite families requires auxiliary measurability arguments.
Finite sums of coordinates are shown to be measurable, predicates of the form
$\exists S \in F.\; \cdots$ are measurable over the product space, and the resulting \texttt{LEAST} expressions preserve measurability.
These facts ensure that all events appearing in the above decomposition are measurable and that the standard independence and union-bound lemmas can be applied without further side conditions.

\paragraph{FP.}

In Isabelle/HOL, the above structural steps are realized by the following core lemmas and theorems:

\medskip

\noindent
\textbf{(a) Coordinate support and point probabilities.}
We first show that for any $x \in U$, the random variable $w(x)$ takes values almost surely in $\{1,\dots,N\}$. This yields the following support lemma:
\begin{lstlisting}[language=Isabelle, caption={Coordinate Support}]
lemma w_pmf_component_in_range:
  assumes "x \in U" "N >= 1" "finite U"
  shows "set_pmf (map_pmf (\lambdaw. w x) (w_pmf N U)) \subseteq {1..N}"
\end{lstlisting}


Based on this, we derive the uniform point probability formula for each $k \in \{1,\dots,N\}$ (appearing as a subgoal in the main proof):
\begin{lstlisting}[language=Isabelle, caption={Point Probability of One Coordinate}]
measure_pmf.prob (w_pmf N U) {w. w x = k} = 1 / real N
\end{lstlisting}

\medskip

\noindent
\textbf{(b) Measurability lemmas (measurability of $\alpha$).}
To legitimately apply probabilistic reasoning, we must show that the auxiliary quantity $\alpha(x)$ is a measurable random variable. The formalization culminates in:
\begin{lstlisting}[language=Isabelle, caption={Measurability of Alpha}]
lemma alpha_measurable:
  assumes "finite U" "F \subseteq Pow U" "x \in U"
  shows "(\lambdaf. alpha (\lambday. if y \in U-{x} then f y else 0) F x)
         \in (PiM (U-{x}) (\lambda_. count_space UNIV)) ->M (count_space UNIV)"
\end{lstlisting}

This proof relies on measurability of finite sums, measurability of finite existential predicates, and the measurability theorem for the \texttt{LEAST} operator.

\medskip

\noindent
\textbf{(c) Independence lemma.}
Next, we lift coordinate-wise independence to random-variable independence by showing that $w(x)$ is independent of $\alpha(x)$, which depends only on the remaining coordinates:
\begin{lstlisting}[language=Isabelle, caption={Independence of Coordinate and Alpha}]
prob_space.indep_var (measure_pmf (w_pmf N U))
  (count_space UNIV) (\lambdaw. w x)
  (count_space UNIV) (\lambdaw. alpha w F x)
\end{lstlisting}

This step combines restriction lemmas for product measures with the previously established lemma \texttt{alpha\_independent\_of\_coord}, ensuring that $\alpha(x)$ depends solely on the weights assigned to $U-\{x\}$.

\medskip

\noindent
\textbf{(d) Main theorem: Isolation Lemma.}
Finally, combining the single-coordinate probability bound with the union bound yields the formalized version of the Isolation Lemma:
\begin{lstlisting}[language=Isabelle, caption={Main Theorem: Isolation Lemma}]
theorem isolation_lemma_main:
  fixes U :: "'a set" and F :: "'a set set" and N :: nat
  assumes "finite U" "F \subseteq Pow U" "F \noteq {}" "N >= 1"
  shows "measure_pmf.prob (w_pmf N U){w. has_unique_minimizer F w}
         >= 1 - real(card U) / real N"
\end{lstlisting}

This theorem completes the formal derivation from the structural collision event to the global lower bound on the probability of uniqueness.

Taken together, these lemmas bridge the gap between the informal probabilistic argument and its fully formal counterpart.
They make explicit the two technical ingredients that are implicit in pen-and-paper proofs: the measurability of $\alpha(x)$ and its independence from the coordinate $w(x)$.
With these in place, the Isolation Lemma follows directly by applying the union bound inside Isabelle/HOL's probability framework.

\section{Chapter Summary}
In this chapter, we presented a fully formal proof of the Isolation Lemma in Isabelle/HOL, by systematically refining a classical proof into structured proof steps and finally machine-checked Isabelle/HOL code.
We reformulated non-uniqueness as the existence of two distinct minimizers, introduced the auxiliary quantity $\alpha(x)$, and showed how it captures the precise collision event responsible for ties.
A key technical ingredient is showing that that $\alpha(x)$ is independent of the individual weight $w(x)$, which enables a reduction to single-coordinate probabilistic events.
Finally, by establishing measurability, independence, and a uniform point probability bound within a product probability space, we derived the standard $1 - |U|/N$ uniqueness bound in a fully mechanized manner.