\chapter{Perfect Matching Problem}
\label{chp:matching}

\section{Introduction to Randomized Parallel Matching}
\label{sec:matching_intro}

The Perfect Matching problem is a fundamental problem in combinatorial optimization and complexity theory.
Formally, given a bipartite graph $G = (U \cup V, E)$ where $|U| = |V| = n$. a \textit{matching} is defined as a set of edges $M \subseteq E$ such that no two edges in $M$ share a common vertex.
A matching $M$ is said to be \textit{perfect} if it contains exactly $n$ edges, effectively covering every vertex in the graph.

While efficient sequential algorithms for finding perfect matchings, such as the augmenting path method, have been well-established, developing an efficient parallel algorithm remains a significant challenge.
This gap between sequential and parallel complexity is not merely a matter of implementation, but is rooted in the structural properties of the problem.

\subsection{Decision versus Construction}
In complexity theory, there is a notable distinction between the \textit{decision} version of a problem (answering if a matching exists) and the \textit{construction} version(outputting the edges of the matching).
For the perfect matching problem, the decision version can be solved in parallel by calculating the determinant of the Tutte matrix, which belongs to the complexity class \textbf{NC}.

However, constructing the matching itself is significantly harder in a parallel setting.
The standard sequential approach of greedily picking edges or finding augmenting paths is inherently serial, as each step depends on the state resulting from the previous step.
This sequential dependency makes it difficult to distribute the workload across multiple processors effectively.

\subsection{The Symmetry Barrier and Coordination}

The primary obstacle to parallelizing the search for a perfect matching is known as the \textit{Symmetry Barrier} \cite{Gupta2011Lec10}.
In many graphs, there can be an exponential number of distinct perfect matchings.
In a sequential setting, an algorithm can easily break this symmetry by picking the first valid matching it finds.

In a parallel environment, multiple processors must work on different edges simultaneously.
If the graph contains multiple perfect matchings, say $M_1$ and $M_2$, one processor might select an edge from $M_1$, while another selects an edge from $M_2$.
Because these edges may not belong to the same global solution, their union is unlikely to form a valid perfect matching.

Without a mechanism to ensure that all processors converge on the \textit{same} solution, parallel coordination becomes impossible.

\subsection{Isolation as a Solution}

To overcome the symmetry barrier, we utilize the \textbf{Randomized NC (RNC)} strategy proposed by Mulmuley, Vazirani, and Vazirani \cite{Gupta2011Lec10}.
The core idea is to transform the symmetric solution space into an asymmetric one by introducing randomness.

By assigning random weights to each edge, we ensure that different perfect matchings likely have different total weights.
The \textbf{Isolation Lemma}, formally proved in Chapter \ref{chp:proof}, provides the mathematical guarantee that with high probability, there is a \textit{unique} minimum-weight perfect matching.

This uniqueness is the foundation of the parallel algorithm: if there is only one target solution, every processor can independently decide if its edge is part of that specific matching.
The need for inter-processor communication is eliminated, and the symmetry is broken.
In the following sections, we formalize how this uniqueness allows us to "extract" the matching using algebraic methods.

\section{Algebraic Modeling: The Tutte Matrix}
\label{sec:tutte_matrix}

To design an efficient parallel algorithm, we must transform the combinatorial search for matchings into a problem that can be solved using parallel linear algebra.
The standard approach utilizes the \textit{Tutte matrix} \cite{mulmuley1987matching} (often referred to as the Edmonds matrix in the bipartite case) to encode the graph structure into a matrix form.

\subsection{Matrix Construction in Isabelle/HOL}

In our formalization, we consider a bipartite graph $G = (V, E)$ with $n$ vertices on each side.
We represent the weight of an edge $(i, j)$ as $w_{i,j}$.
In the \texttt{bipartite\_graph} locale, the matrix $D$ is constructed such that each entry $D_{i,j}$ encodes both the existence of an edge and its weight:

\begin{lstlisting}[language=Isabelle, caption={Matrix Entry Definition}]
definition D_entry :: 
    "nat => nat => int" where
    "D_entry i j = (if (i,j) : E then (2::int) ^ (w i j) else 0)"
\end{lstlisting}

The choice of $2^{w_{i,j}}$ as the matrix entry, as proposed in \cite{mulmuley1987matching, Gupta2011Lec10} is a critical design decision to prevent algebraic cancellation.
By using powers of 2, the total weight of a perfect matching becomes the exponent in the Leibniz expansion of the determinant.
This algebraic encoding allows the determinant to "store" information about all feasible matchings simultaneously within its binary representation.

\subsection{Determinant and Perfect Matchings}

A perfect matching in a bipartite graph can be formally defined as a permutation $p$ of the set $\{0, \dots, n-1\}$ such that every pair $(i, p(i))$ is a valid edge in $E$.
Our formal definition in Isabelle/HOL captures this requirement:

\begin{lstlisting}[language=Isabelle, caption={Perfect Matching Definition}]
definition is_perfect_matching :: 
    "(nat => nat) => bool" where
  "is_perfect_matching p <-> 
    (p permutes {0..<n} & (ALL i < n. (i, p i) : E))"
\end{lstlisting}

The determinant of matrix $D$, defined via the Leibniz formula, serves as the primary tool for isolation:

\begin{equation}
    det(D) = \sum_{p \in S_n} \text{sgn}(p) 
    \prod_{i=0}^{n-1} D_{i, p(i)}
\end{equation}

By substituting the definition of $D_{i,j}$, any term in the sum is non-zero if and only if the permutation $p$ corresponds to a valid perfect matching.
For such a permutation, the contribution to the determinant is:

\begin{equation}
    \text{sgn}(p) \cdot \prod_{i=0}^{n-1} 2^{w_{i, p(i)}} 
    = \text{sgn}(p) \cdot 2^{\sum w_{i, p(i)}} 
    = \text{sgn}(p) \cdot 2^{weight(p)}
\end{equation}

If $p$ does not represent a perfect matching, at least one $D_{i, p(i)}$ is 0, causing the entire product to vanish.
This mapping ensures that the determinant is a weighted sum of all perfect matchings, providing the necessary algebraic foundation for the subsequent extraction of the unique minimum matching.

\section{The Weight-to-Multiplicity Bridge}
\label{sec:weight_multiplicity}

The core of the parallel matching algorithm lies in the relationship between the determinant of the Tutte matrix and the weights of the perfect matchings it represents.
Once the Isolation Lemma ensures that the minimum-weight perfect matching $p_{min}$ is unique, the determinant $det(D)$ effectively becomes a binary encoding of the matching weights.

\subsection{The Non-Cancellation Principle}
In the Leibniz expansion of $det(D)$, each perfect matching $p$ contributes a term of the form $\text{sgn}(p) \cdot 2^{weight(p)}$.
A potential concern in this algebraic sum is that terms with opposite signs might cancel each other out.
However, because the weights are stored as exponents of 2, the term with the smallest exponent cannot be cancelled by any combination of terms with larger exponents.

Formally, if $W_0$ is the unique minimum weight, $det(D)$ can be expressed as:
\begin{equation}
    det(D) = \text{sgn}(p_{min}) 2^{W_0} + 
    \sum_{p \in P \setminus \{p_{min}\}} \text{sgn}(p) 2^{weight(p)}
\end{equation}
Since for all other matchings $p$, $weight(p) > W_0$, every term in the summation is a multiple of $2^{W_0+1}$.
Therefore, the entire determinant is a multiple of $2^{W_0}$ but not of $2^{W_0+1}$, ensuring that the lowest non-zero bit of the sum is preserved.

\subsection{Formalization via Multiplicity}
To extract this minimum weight in Isabelle/HOL, we utilize the \texttt{multiplicity} function from the number theory library.
The multiplicity of a prime $k$ in an integer $x$ is the largest power of $k$ that divides $x$.
In our binary context, \texttt{multiplicity 2 det\_D} yields exactly $W_0$.

We formalized this connection through several key lemmas.
First, \texttt{multiplicity\_2\_abs} establishes the base case for powers of 2:

\begin{lstlisting}[language=Isabelle, caption={Multiplicity of Power of 2}]
lemma multiplicity_2_abs:
  fixes x :: int
  assumes "abs x = 2 ^ k"
  shows "multiplicity 2 x = k"
\end{lstlisting}

Then, we established the main theorem relating the determinant to the unique minimum weight:

\begin{lstlisting}[language=Isabelle, caption={Multiplicity of Determinant}]
lemma unique_det_power_of_weight:
  assumes "unique_min_weight_condition"
  shows "multiplicity 2 det_D = min_weight_val"
\end{lstlisting}

This result is proven in Isabelle by applying \texttt{multiplicity\_eqI}, which requires showing that $2^{W_0}$ divides the sum and $2^{W_0+1}$ does not.
The proof handles the summation by splitting it into the unique minimum term and the remaining "heavy" terms, formalizing the intuitive non-cancellation argument presented earlier.

\section{Parallel Edge Extraction Logic}
\label{sec:edge_extraction}

The final stage of the parallel algorithm is to identify which edges $(i, j) \in E$ constitute the unique minimum-weight perfect matching $p_{min}$.
Crucially, this step must be performed in parallel, meaning each edge must be testable independently of others.

\subsection{Minors and Subgraph Matchings}
For any edge $(i, j)$ we consider the minor $D_{i,j}$, which is the submatrix obtained by deleting the $i$-th row and $j$-th column of $D$.
Algebraically, the determinant of this minor represents the sum of all perfect matchings in the subgraph $G \setminus \{u_i, v_j\}$.
If an edge $(i, j)$ belongs to a perfect matching $p$ in $G$, then $p \setminus \{(i, j)\}$ must be a perfect matching in the reduced graph.

In our formalization, we define \texttt{minor\_sum i j} to represent the contribution of all matchings containing the edge $(i, j)$.
This relationship is captured by the lemma \texttt{minor\_sum\_decomposition}:

\begin{lstlisting}[language=Isabelle, caption={Minor Sum Decomposition}]
lemma minor_sum_decomposition:
  assumes "i < n" "j < n" "(i,j) : E"
  shows "minor_sum i j = D_entry i j * cofactor_sum i j"
\end{lstlisting}

This decomposition allows us to isolate the contribution of a single edge weight from the rest of the matching structure.

\subsection{The Parallel Decision Predicate}
To determine if $(i, j) \in p_{min}$, we compare the minimum weight of matchings containing $(i, j)$ with the global minimum weight $W_0$.
If $(i, j)$ is part of the unique minimum matching, the minimum weight in the subgraph plus the edge weight $w_{i,j}$ must equal the global minimum weight.

This core division logic can be expressed by the following identity:
\begin{equation}
    w(i,j) + \text{multiplicity}_2(\text{det}(D_{i,j})) = min\_weight\_val
\end{equation}

We formalized this as a necessary and sufficient condition in the theorem \texttt{edge\_test\_iff\_correct}:

\begin{lstlisting}[language=Isabelle, caption={Correctness of Edge Extraction}]
theorem edge_test_iff_correct:
  assumes p_min_def: 
    "is_perfect_matching p_min" 
    "matching_weight p_min = min_weight_val"
  assumes p_min_unique: 
    "ALL q. is_perfect_matching q & matching_weight q 
            = min_weight_val --> q = p_min"
  shows "p_min i = j <-> 
         (cofactor_sum i j ~= 0 & w i j + multiplicity 2 (cofactor_sum i j) = min_weight_val)"
\end{lstlisting}

This theorem provides the formal justification for the parallel algorithm.
Each edge $(i, j)$ can be tested independently by computing the determinant of its corresponding minor.
If the weight condition holds, the edge is guaranteed to be part of the unique matching $p_{min}$.
Since determinant calculation is in \textbf{NC}, the entire matching can be constructed in parallel time.

\section{Probabilistic Correctness via Isolation Lemma}
\label{sec:matching_prob_correctness}

The formal verification of the parallel matching algorithm is completed by establishing its probabilistic success guarantee.
While Chapter \ref{chp:proof} provides a general proof of the Isolation Lemma for any finite family of sets, we must now specialize this result to the domain of perfect matchings in a bipartite graph.

\subsection{Mapping Graphs to Sets}
To apply the Isolation Lemma, we first define a formal mapping between graph-theoretic structures and the set-theoretic framework.
The universe $U$ is identified with the set of edges $E$, and the family $\mathcal{F}$ is defined as the collection of all edge sets that constitute a valid perfect matching:

\begin{lstlisting}[language=Isabelle, caption={Perfect Matching Family}]
definition perfect_matching_family :: 
    "nat => (nat * nat) set => ((nat * nat) set) set" where
    "perfect_matching_family n E = 
      {matching_to_set n p | p. 
       p permutes {0..<n} & (ALL i < n. (i, p i) : E)}"
\end{lstlisting}

A critical step in this bridge is ensuring that the weight of a matching in the graph model is identical to the weight of the corresponding set in the isolation model.
We established this via the lemma \texttt{weight\_equivalence}.
Furthermore, the lemma \texttt{matching\_to\_set\_inj} ensures that each permutation corresponds to a unique set of edges, preventing any unintended collisions in the probability space.

\subsection{Final Probability Bound}
The equivalence between the graph-theoretic unique minimum condition and the set-theoretic \texttt{has\_unique\_minimizer} predicate is established in the lemma \texttt{pm\_iso\_equivalence}.
This allows us to lift the general isolation bound to the specific case of bipartite matching.

The theorem \texttt{isolation\_lemma\_for\_perfect\_matching} provides the machine-checked lower bound on the success probability:

\begin{lstlisting}[language=Isabelle, 
    caption={Isolation Lemma for Perfect Matching}]
theorem isolation_lemma_for_perfect_matching:
  fixes N :: nat
  assumes "N >= 1"
  assumes "EX p. is_perfect_matching p"
  assumes "E <= {0..<n} * {0..<n}"
  shows "measure_pmf.prob (w_pmf N E)
      {w. unique_min_weight_condition (\lambda u v. w (u,v)) E}
      >= 1 - (real (card E) / real N)"
\end{lstlisting}

This theorem completes the formal derivation.
For a graph with edge set $E$ and weights chosen uniformly from $\{1, \dots, N\}$, the probability that the algorithm identifies a unique matching is at least $1 - |E|/N$.
This result provides the formal justification for using randomized weighting to achieve parallel coordination in the Perfect Matching problem.

\section{Formalization Challenges in Isabelle/HOL}
\label{sec:matching_impl}

The formalization of randomized parallel matching involves several technical challenges that are absent in informal presentations.
These difficulties arise primarily from the strict requirement for type safety and the integration of disparate mathematical libraries within Isabelle/HOL.

\subsection{Bridging Type Systems: Int versus Nat}

A significant hurdle during implementation was the reconciliation of different numeric types.
In Isabelle/HOL, the \texttt{multiplicity} function is defined over the \texttt{int} type within the number theory library.
However, the \texttt{Determinants} library often operates over more general algebraic structures.

Since the edge weights are naturally non-negative, we initially modeled them as \texttt{nat}.
Yet, the determinant of a matrix with \texttt{sign p} (which can be $-1$) requires the codomain to be \texttt{int}.
This necessitated the use of explicit type casting and lemmas to prove that properties like multiplicity are preserved.
This technical detail required formal justification via lemmas such as \texttt{multiplicity\_2\_not\_dvd\_Suc}.

\subsection{Library Integration and Automation}

Another challenge involved the integration of the \texttt{Permutations} library with the \texttt{Probability} library.
To prove the final probability bound, we had to ensure that the combinatorial structure could be measured within the product probability space.

A major effort was spent on the lemma \texttt{unique\_subset\_sum\_power\_of\_weight}.
In an informal setting, one might simply state that higher powers of 2 do not affect the lowest bit.
In Isabelle, this requires a rigorous proof to show that a sum of multiples of $2^{k+1}$ remains a multiple of $2^{k+1}$, thus not interfering with a single term of value $2^k$.
The use of structured proof steps allowed us to manage this complexity by breaking the summation logic into manageable sub-goals. 

\section{Chapter Summary}
\label{sec:matching_summary}

In this chapter, we have presented a complete formal verification of a randomized parallel algorithm for the Perfect Matching problem.
By constructing the problem as a bipartite graph within Isabelle/HOL, we demonstrated how the abstract probabilistic guarantees of the Isolation Lemma can be translated into concrete algebraic properties of the Tutte matrix.

We formally established that by encoding edge weights as powers of 2, the determinant of the matrix becomes a structured representation of all possible matchings.
The core of our verification relied on the "Weight-to-Multiplicity" bridge, where we proved that the multiplicity of 2 in the determinant identifies the exact weight of the unique minimum-weight matching.
Furthermore, we verified the parallel edge extraction logic, proving that each edge can independently determine its membership in the matching by calculating its corresponding minor.
This chapter completes the transition from the foundational probability theory established in Chapter \ref{chp:proof} to a high-level application in randomized algorithm analysis, achieving a fully machine-checked proof of correctness for a Randomized NC (RNC) algorithm algorithm.
